%!TEX root = gutter+stars.tex
\chapter*{Acknowledgements}

This dissertation was completed due to the support of many people.
I would like to express my thanks to all of you, being aware of the fact that this enumeration might be incomplete in both its quantity and quality.

\begin{center}* * *\end{center}

First and foremost my thank goes to Oscar Nierstrasz. 
%
I must thank him not only -- like many have done before me -- for the opportunity to work at the Software Composition Group, but much more for setting a unique example and role model as researcher. For the dedication and care he committed to me and my 
work, to the lectures he holds and to all of his students in general. It is this disposition that 
nurtured my motivation and interest for the field of computer science since the first lecture 
I had with him. 

My thank goes to Harald Gall for accepting to be on my PhD committee and for the inspiring discussions on my visits and talks in Zurich. 

My thank goes to Horst Bunke for accepting to chair my PhD defense and for his help and advice when I was stuck with the layout of software maps.

Special thank goes to Rob DeLine for hosting and mentoring me as an intern at Microsoft research. Rob's group is not only doing great work, but is also an awesome group of people to work with. Andrew Begel, Kael Rowan and Gina Veniola --- it was a such great and unique time, thank you!

Special thank goes to Sushil Bajracharya and Yunwen Ye for co-organizing the SUITE workshop. For sharing the vision of a true working workshop and for all their dedication and hard work across three-time zones. Thank you!

Most special thank goes to my students David Erni, Manuel Friedli, Florian Gysin, Lea H\"ansenberger, Peter Loretan and Dominique Matter for their trust and dedication. For the countless hours spent in the lab, for all the deadline-frenzy night shifts, and for the many weizen beer at the Cinebar. It was a pleasure working with you guys, you're awesome!

Special thank also goes to my friends and peers Markus Gaelli, Niko Schwarz, Toon Verwaest, and Erwann Wernli. For the great many discussions and hands-on hacking sessions in the lab, for all your inspiration and support, and for the many evenings out in town.  Thank you guys!

My thank also goes to Marcus Denker for being a dear friend and for mentoring me when the group lacked a proper postdoc. For sharing his experience and advice, for his thoughtful insight into research and the world in general, and for all the awesome book recommendations. 

My thank also goes to Tudor Girba for his supervision during my Master's thesis and later, as a peer, for the opportunity for personal growth.

My thank goes to all peers and co-authers, both internal and external, whom I forgot to name here. For sharing your experience, for the great many discussions and for all you're inspring work. You know whom you are, thank you!

My thank also goes to Aaron Funk of Venetian Snares for producing the music that kept my awake when working late night towards a deadline. 

Special thank goes to my followers on Twitter and to the anonymous readers of my blog. For sharing and caring. You're feedback and interest in my work is great motivation to me, thank you guys!

Special thank also goes to the internet for making all people equal, for bringing lolcats and freedom to the world, and for just being awesome.

Eventually, my thank goes to Nadine Michel, the mother of my dear son Felix, for her love and care. 

\begin{center}* * *\end{center}

Thank you all!
